Grow with the flow: faster growth is associated with more variable precipitation in a perennial herb

### 2016-01-20 ms/ms.tex
Lots of changes.

\documentclass[11pt, oneside]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}

@@

%
% Start document
%
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\maketitle

$^1$ Biodiversity Research Centre, University of British Columbia, Vancouver, BC, Canada

$^*$corresponding author: Chris Muir, cdmuir@biodiversity.ubc.ca

Original Article

Key words: local adaptation, photosynthesis, growth rate, climatic variance, \textit{Mimulus}

Word counts: TBA

3 figures; 3 tables; ? supporting files

Data will be archived on Dryad upon acceptance.

@@

Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it. Local adaptation within species most often involves genetic differences in ecologically relevant traits caused by selection to different environments, leading to predictable trait-environment correlations that persist in a common environment. Although local adaptation is well-supported by data from reciprocal transplant experiments, trait-environment associations, and ecological genetics/genomics, predicting which traits respond to selection is beyond our current understanding. In particular, the role of physiological traits is poorly understood, even though physiology explains adaptation to the abiotic environment, which we need to understand as a baseline for anticipating how organisms will respond to climate change. A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures. Here, we  investigate which traits and causal selective agents explain range-wide patterns of local physiological adaptation to climate in \textit{Mimulus}, a model genus for local adaptation studies.

% niche-tracking versus niche-shifting
One surprising pattern is that populations adapt to different environments by adjusting their life history to stay within the same fundamental physiological niche rather than have the niche itself evolve. For example, mosquitoes adjust diapause length with latitude rather than evolve altered ...\cdm{what other examples could we include here?} This suggests that in many situations it is evolutionarily easier to change life history rather than fundamental physiological tolerance.

% Selective agents
We also know little about the selective agents, such as climatic factors or biotic interactions, that are most important in driving adaptive divergence. In certain situations, such as pollinator shifts and edaphic adaptation, it is straightforward to identify putative selective agents and test predicted mechanisms. However, it is more challenging for cases in which traits vary with less specific factors like mean annual precipitation or latitude. For example, body size clines (larger size at higher latitude) have evolved repeatedly on three contients in \textit{Drosophila subobscura}, strongly indicating selection \citep{Huey_etal_2000}. But which one or more of the many climatic factors that covary with latitude actually controls selection on body size? Selection might be even more complex for perennial species that live over multiple seasons, meaning that environmental variation rather than the average environment may be a selective agent.

% Motivate using Mimulus
Linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions. Many classic and contemporary studies of local adaptation have been conducted using species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources [willis, lowry?]. Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008}.

Two common signatures of local adaptation are nonrandom genetic differences in ecologically important traits and variation in plasticity (genotype by environment interaction). In this study, we found strong genetically-based differences in photosynthesis and growth, but no evidence for variation in plasticity in response to temperature and drought treatments. Interannual variation in precipitaiton is associated intrinsic variation in these traits, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M. cardinalis}.

@@

We used 16 populations from throughout the range of \textit{M. cardinalis} (Table~\ref{table:Table_FocalPops}). Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.

@@

\begin{table}[ht]
   \centering
   \topcaption[Focal Populations]{Geographic region, latitude, longitude, and elevation (mas = meters above seal level) of 16 focal populations used in this study.}
   \begin{tabular}{@{} lllllll @{}}
      \toprule
  Name& Region  & Demo?& Pop gen?& Latitude  & Longtiude  & Elevation (mas) \\
      \midrule
	HAU & South Margin		& yes	& yes	& 32.657	& 
    \ensuremath{-116.532}	& 799   \\
	CTC	& South Margin		& yes	& no	& 32.609 & 
    \ensuremath{-116.7}	& 267   \\
	CUR	& South Margin		& yes	& yes & 32.9 & 
    \ensuremath{-116.585}	& 1180   \\
	GRP & South Margin		& no	& no	& 33.314 &
    \ensuremath{-116.871}	& 1577   \\
	WWC &	Transverse		  & yes	& yes	& 33.994 & 
    \ensuremath{-116.665}	& 705   \\
	MIL	& Transverse	  	& yes	& no	& 34.077 & 
    \ensuremath{-116.873}	& 2050   \\
	WFM	& Transverse	  	& yes	& no	& 34.284 & 
    \ensuremath{-117.378}	& 1120   \\
	NMT	& South Sierras		& yes	& yes?& 36.201 & 
    \ensuremath{-118.651}	& 1314   \\
	PRD	& South Sierras		& yes	& yes	& 36.518 & 
    \ensuremath{-118.759}	& 926   \\
	RWD	& South Sierras		& yes	& yes	& 36.691 & 
    \ensuremath{-118.91}	& 1727   \\
	WNA	& Central Sierras	& yes	& yes	& 37.541 & 
    \ensuremath{-119.649}	& 1224   \\
	RBW	& Central Sierras	& yes	& no	& 37.819 & 
    \ensuremath{-120.007}	& 876   \\
	MYU	& North Sierras		& yes	& yes	& 39.397 & 
    \ensuremath{-121.082}	& 455   \\
	LIJ	& North Sierras		& yes	& yes	& 39.743 & 
    \ensuremath{-120.704}	& 1603   \\
	DPC	& North Coast	  	& yes	& yes	& 41.668 & 
    \ensuremath{-123.11}	& 707   \\
	RCC	& North Margin		& yes	& yes	& 43.374 & 
    \ensuremath{-122.957}	& 326   \\
	\bottomrule
	\end{tabular}
	\label{table:Table_FocalPops}
\end{table}

@@

We imposed four treatments, a fully-factorial cross of two temperature treatments and two watering treatments. Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment and another to the Cool treatment. Within each chamber, there were two Well-watered blocks and two Drought blocks. A detailed description of treatments is given below \cdm{should some be moved to supplement?} and summarized in Fig~\ref{fig:Fig_ExptlDes}. The irradiance in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$. The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature treatments (data not shown).

\subsubsection*{Temperature treatments}

We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (Hot) and Little Jameson (Cool). We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}. Daily temperatures from ClimateWNA are highly correlated with the air temperature recorded from data loggers in the field at these sites (A. Angert, unpub. data). Hence, the ClimateWNA temperature profiles are similar to actual thermal regimes experienced by \textit{M. cardinalis} in nature. We simulated realistic temperature regimes by calculating the mean temperature trend from June to August using LOESS \citep{Cleveland_etal_1992}. The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ). The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals. We fit and simulated from the VAR(2) model using the package \pkg{dse} \citep{Gilbert_2014} in R. Simulated data closely resembled the autocorrelation and between-site correlation of the actual data. From simulated mean temperature, we next selected minimum and maximum daily temperatures. Mean, min, and max temperature were highly correlated at both sites. We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models. For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature. During the middle of the day, temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.

\subsubsection*{Watering treatments}

For watering treatments, we simulated two extreme types of streams where \textit{M. cardinalis} grows. In the well-watered treatment, we simulated a large stream that never goes dry during the summer growing season. In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season, but gradually dries down as the winter snow pack melts. In both treatments, plants were bottom-watered using  water chilled to 7.5 [check] by MAKE AND MODEL OF CHILLER. Plants in the well-watered treatment were fully saturated every two hours during the day. Watering in the drought treatment gradually declined from every two hours to every day between May 20 (36 days after sowing) and 10 June (57 days after sowing). Simultaneously, the amount of bottom-watering per flood decreased, such that only the bottom of the cone-tainers were wetted by the end of the experiment.

\cdm{Are there any data from California about soil water content through the season?}

@@

We censused longest leaf length $>1$ mm on 10 days (twice per week) between 12 May and 12 June (28 to 59 days after sowing). We ceased measuring leaf length once it appeared to asymptote and growth shifted to shoot elongation.  We also censused plant height on 7 days (twice per week) between 29 May and 20 June (45 to 67 days after sowing). Both leaf expansion and shoot elongation were modeled as a second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.

\paragraph{Photosynthesis}
During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of 329 plants evenly spread between treatments and families within populations. The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska). All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), temperature, and moderate relative humidity. During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ \cdm{get exact temps used}) so that all plants within a temperature treatment were measured under the same conditions.

@@

\paragraph{Biomass (show correlation between growth rate and biomass?)}

@@

\subsection*{Principal components of germination, growth, and photosynthesis}
We summarized population-level coefficients, after factoring out Treatment and other effects, using principal components. The first principal component of these traits (Trait PC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast and slow growing populations.

@@

As with previous studies in \textit{Mimulus} and many other species [CITES, \citep{Huey_etal}, mosquitoes, what else?], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g. temperature or precipitation) that covary with latitude. We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We did this by looking for overlap between climatic variables that best predict latitude of \textit{M. cadinalis} occurence records and climatic variables that best predict trait variation across our 16 focal populations. For brevity, we refer to these as Climate-Latitude and Climate-Trait variables. We selected Climate-Latitude and Climate-Trait variables independently using Random Forest (VSURF algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}). From VSURF models, we kept only variables selected for prediction, the most stringent criteria.

For these analyses, we compiled a representative set of 178 recent (since 2000) known \textit{M. cardinalis} occurences. These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included 24 cliamte variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}). For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation. Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.

@@

Genetic differences between populations affecting originsmal function and performance point to traits that may be involved in local adaptation. Using a common garden design, we identified strong genetic differences in germination, growth, and photosynthetic rate among populations of \textit{M. cardinalis} (Table~\ref{table:Table_AnovaSummary}). A single principal component captured 74.2 %$\%$ of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}). As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.

@@

\begin{figure}[h!]
	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_PC1vLat.pdf}}
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[Southern populations grow faster]{Trait variation, from fast to slow growth, is closely associated with latitude. Each point is a population, its latitude of origin a position along the slow to fast growth axis, defined as Principal Component 1 of five traits (see Methods). The line and 95\% confidence intervals (grey polygon) were estimated using standard linear regression.}
	\label{fig:Fig_PC1vLat}
\end{figure}

@@

Genotype $\times$ environment interactions are also a common signature of local adaptation [CITE]. For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions. We found remarkably little evidence for this pattern in \textit{M. cardinalis}. There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing. Otherwise, populations responded remarkably similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}). Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects. However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.

@@

@@

\bibliography{refs}
\bibliographystyle{evolution}

@@

	\bottomrule
	\end{tabular}
	\end{center}
	\label{tables:TableS_InitialSize}
\end{table}

@@

@@

\subsection*{Spatially-averaged climate}

>>>>

% TO DO:
% 1. Is water regime on figure 1 correct?
% 2. Finalize Table 2
% 3. Do anything with biomass data?
% 4. Citations for latitudinal clines
	% cases where clinal variation has 'obvious' selective source:
	% Antonovics and Bradshaw discuss steepness of clines and strength of selection over 10's of meters off mines in Anthoxanthum
	% Caicedo et al 2004?
	% clines in phenotypic traits reviewed by Hedrick 2006, Schemske and Bier 2007
% 5. Citations for GxE as signature of local adaptation
% 6. Fix figure 2
% 7. Supplemental Figure and Table captions
% 8. Supplemental figure showing agreement between assigned to climatic variables from point estimated and spatially averaged analyses

\documentclass[11pt, oneside]{article}

@@

%
% Start document
%

\usepackage{Sweave}
\begin{document}
\input{ms-concordance}
\maketitle

$^1$ Biodiversity Research Centre, University of British Columbia, Vancouver, BC, Canada \\
$^*$corresponding author: Chris Muir, cdmuir@biodiversity.ubc.ca \\

Original Article \\

Key words: local adaptation, cline, photosynthesis, growth rate, \textit{Mimulus} \\

Word counts: TBA \\
3 figures; 3 tables; ? supporting files \\
Data will be archived on Dryad upon acceptance. \\

\cdm{Chris' comments}\\
\ala{Amy's comments}

@@

% Local adaptation and clines
Local adaptation within species is ubiquitous; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Hereford_2009}. Local adaptation also frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977}. Phenotypic differences between populations along the cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}. Despite a long history of studying local adaptation and clines, it remains to challegenging to identify exactly which traits are under selection and which differ for nonadaptive reasons. In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment. We need to understand physiological adaptations within species as a baseline for anticipating how organisms will respond to climate change. A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures. Here, we examine physiological trait variation and possible selective agents in the perennail herb \textit{Mimulus cardinalis} Douglas ex Benth. (Phrymaceae), a model system for local adaptation studies.

% Intrinsic versus plastic variation
There are two basic approaches one can use to identify candidate traits underlying local adaptation in a common garden. First, if genetically-based trait differences between populations vary clinally with environmental differences, this may point to traits important for local adaptation. Second, genotype by environment interactions could indicate that variation in plasticity mediates local adaptation. We distinguish between these signatures of local adaptation by referring to `intrinsic' and `plastic' trait variation, respectively. There are classic cases adaptation involving both intrinsic and plastic trait variation. For example, intrinsic differences in critical photoperiod [CITE] and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment. Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype by environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.

% Selective agents
Either instrinsic and/or plastic variation should vary clinally along environmental gradients. Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown. For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}. Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation. Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely related to climatic variation. Despite the fact that latitudinal clines in particular have been studied for a long time, the exact climatic factors (e.g. minimum or maximum temperature, growing season length, strength of biotic interactions) is rarely known. Because many climatic factors vary latitudinally, and which climatic factors vary latitudinally changes over the earth's surface (e.g. coastal vs. continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}. In plants especially, we know little about the prevelance and adaptive significance of clinal variation in fundamental physiological traits like photosynthesis and the impact on plant performance.

% For the most part, we know much more about selective agents responsible for clines for conspicuous things like colouration, such as in mice \citep{Sumner_1932, Mullen_Hoekstra_2008}

% Motivate primary questions using Mimulus
In this study, we address these gaps by asking whether intrinsic or plastic physiological trait variation corresponds with latitude and what climatic factor(s) could plausibly be repsonsible for latitudinal clines in a focal species, \textit{Mimulus cardinalis}. We chose this species because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions. Many classic and contemporary studies of local adaptation use species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}. Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010}. We measured genetic and genotype by environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude. We found a latitudinal cline of intrinsic differences in photosynthesis and growth, but no evidence for variation in plasticity. Interannual variation in precipitaiton and/or precipitation seasonality is associated this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M. cardinalis}. We place these findings in the context of life history theory and consider future directions in the Discussion.

@@

We used 16 populations from throughout the range of \textit{M. cardinalis} (Table~\ref{table:Table_FocalPops}). Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature.

@@

\begin{table}[ht]
   \centering
   \topcaption[Focal Populations]{Geographic region, latitude, longitude, and elevation (mas = meters above seal level) of 16 focal populations used in this study.}
   \begin{tabular}{@{} lllll @{}}
      \toprule
  Name& Region  & Latitude  & Longtiude  & Elevation (mas) \\
      \midrule
	HAU & South Margin & 32.657	& 
    -116.532	& 799   \\
	CTC	& South Margin & 32.609 & 
    -116.7	& 267   \\
	CUR	& South Margin & 32.9 & 
    -116.585	& 1180   \\
	GRP & South Margin & 33.314 &
    -116.871	& 1577   \\
	WWC &	Transverse & 33.994 & 
    -116.665	& 705   \\
	MIL	& Transverse & 34.077 & 
    -116.873	& 2050   \\
	WFM	& Transverse & 34.284 & 
    -117.378	& 1120   \\
	NMT	& South Sierras & 36.201 & 
    -118.651	& 1314   \\
	PRD	& South Sierras & 36.518 & 
    -118.759	& 926   \\
	RWD	& South Sierras & 36.691 & 
    -118.91	& 1727   \\
	WNA	& Central Sierras & 37.541 & 
    -119.649	& 1224   \\
	RBW	& Central Sierras	& 37.819 & 
    -120.007	& 876   \\
	MYU	& North Sierras	& 39.397 & 
    -121.082	& 455   \\
	LIJ	& North Sierras	& 39.743 & 
    -120.704	& 1603   \\
	DPC	& North Coast & 41.668 & 
    -123.11	& 707   \\
	RCC	& North Margin & 43.374 & 
    -122.957	& 326   \\
	\bottomrule
	\end{tabular}
	\label{table:Table_FocalPops}
\end{table}

@@

We imposed four treatments, a fully-factorial cross of two temperature levels and two watering levels. The temperature levels closely simulated an average growing season at the thermal extremes of the species range, which designate as Hot and Cool treatments. Watering levels contrasted a perennial and seasonal stream, which we refer to as Well-watered and Drought treatments. A detailed description of treatments is described in the Supplemental Information and summarized in Fig~\ref{fig:Fig_ExptlDes}. Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment level and another to the Cool treatment level. Within each chamber, there were two Well-watered blocks and two Drought blocks.  The irradiance in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$. The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown).

@@

We censused leaf length twice per week from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements. We ceased measuring leaf length once it appeared to asymptote and growth shifted to shoot elongation.  We also censused plant height on 7 days (twice per week) between 29 May and 20 June (45 to 67 days after sowing). Both leaf expansion and shoot elongation were modeled as a second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.

\paragraph{Photosynthesis}
During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of 329 plants evenly spread between treatments and families within populations. The youngest, fully-expanded leaf \cdm{this is what I did...not sure exactly which node, as I didn't record that when I did measurements} acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska). All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), temperature, and moderate relative humidity. During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level were measured under the same conditions.

@@

\paragraph{Biomass at harvest}

\cdm{show correlation between growth rate and biomass?}

@@

\subsection*{Principal components of germination, growth, and photosynthesis}
For each single-trait model above, we extracted the Population coefficient (factoring out Treatment and other effects). The multivariate distribution of these coefficients was then summarized using principal components analysis (PCA). The first principal component of these traits (Trait PC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast and slow growing populations.

@@

We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results). Since latitude \textit{per se} cannot be a selective agent, we considered three possible explanations. First, latitude may be strongly correlated, in this species, with one or two climatic variables, such as temperature, precipitation, or growing degree-days. Second, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latiduinal cline. Third, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration. For example, average temperature of \textit{M. cardinalis} populations at a given latitude varies widely, but in aggregate, a Southern metapopulation experiences warmer climate than a Northern one. Thus, any particular Southern population would be warm-adapted, even if it was located in cooler (e.g. high elevation) site.

To evaluate support for these hypotheses, we used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We looked for overlap between climatic variables that best predict latitude of \textit{M. cadinalis} occurence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations. For brevity, we refer to these as Climate-Latitude and Climate-Trait variables. We selected Climate-Latitude and Climate-Trait variables independently using Random Forest (VSURF algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}). From VSURF models, we kept only variables selected for prediction, the most stringent criterion.

The first hypothesis predicts that there should be one or two Climate-Latitude and Climate-Trait associations that are strongly correlated, whereas the second hypothesis predicts that several climatic variables should be weakly correlated. To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-Trait variables as above, except that we used spatially averaged climate variables. We sampled climate at 1000 random points (at 90-m resolition) within a 100-km buffer around focal populations. We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}. Since \textit{M. cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}. Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}. For clarity, we distinguish between analyses where climate is inferred from a single point (`point estimated Climate-Trait') versus averaged across a 100-km buffer (`spatially averaged Climate-Trait').

For these analyses, we compiled a representative set of 178 recent (since 2000) known \textit{M. cardinalis} occurences. These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included 24 climate variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}). The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range. For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation. Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.

@@

Using a common garden, we identified strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M. cardinalis}, as evidenced by large and highly significant population effects (Table~\ref{table:Table_AnovaSummary}). A single principal component captured 74.2 \% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}). As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.

@@

\begin{figure}[h!]
	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_PC1vLat.pdf}}
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[Southern populations grow faster]{Trait variation, from fast to slow growth, is closely associated with latitude. Each point is a population, its latitude of origin a position along the slow to fast growth axis, defined as Principal Component 1 of five traits (see Methods). The line and 95\% confidence intervals were estimated using standard linear regression.}
	\label{fig:Fig_PC1vLat}
\end{figure}

@@

Genotype $\times$ environment interactions are also a common signature of local adaptation. For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions. We found remarkably little evidence for this pattern in \textit{M. cardinalis}. There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing. Otherwise, populations responded remarkably similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}). Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects. However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.

@@

% perhaps responses of Simiolus sp (annual, drought selects for escape rather than tolerance) differ from card (perennial, escape through growth rather than flowering time) because drying out of rivers is less severe than seeps and springs?

@@

\bibliography{../../CommonResources/refs}
\bibliographystyle{evolution}

@@

	\bottomrule
	\end{tabular}
	\end{center}
	\label{table:TableS_InitialSize}
\end{table}

@@

% Table of Climate-Latitude and Climate-Trait variables

\begin{table}[ht]
   \centering
   \topcaption[Important climatic variables]{Important climatic variables predicting latitude of \textit{M. cardinalis} populations (`Climate-Latitude') and the first principal component of traits measured in a common garden (`Climate-Trait'). Importance and significance were determined using the variable selection using random forests (VSURF) algorithm (see `Methods'). Climatic variables are described in Table~\ref{table:TableS_ClimVars}. $\mu$ signifies the mean of the climate variables from 1981--2010; $\sigma$ indicates coeffiecient of variation among years}
   \resizebox{6.5in}{!}{
   \begin{tabular}{@{} lll @{}}
      \toprule
    	Climate-Latitude variables & Climate-Trait variables & \\
                            	   & Point estimated         & Spatially averaged \\
      \midrule
			\input{Tables/TableS_ClimVarImp.txt}
			\bottomrule
   \end{tabular} }
   \label{table:TableS_ClimVarImp}
\end{table}

@@

\section*{}

\subsection*{Temperature treatments}
